---
title: Spring 2026 PUBH 422 Project 2 Group 6
output: 
  word_document: default    (!!!!!!REKNIT WHEN EVERYONE IS DONE WITH INTERPRETATIONS!!!!)
  html_document: default
  pdf_document: default
date: "2026-04-17"
---

Group 6 Members: Abril Zarate, Andrew Carias

```{r}
# Task 2:Perform an initial exploration}
library(tidyverse)

NHIS_Data_2021 <- read_csv("data/NHIS _Data_2021.csv")

str(NHIS_Data_2021)
summary(NHIS_Data_2021)
head(NHIS_Data_2021)
```
Interpretation:



```{r}
# Day 2: Task 1 - Variable Selection & Missing Value Check}
  
library(tidyverse)

NHIS_Data_2021 <- read_csv("data/NHIS _Data_2021.csv")

# Select only variables asked for
data_import_R_02_subset <- NHIS_Data_2021 |>
  select(
    AGEP_A,
    WEIGHTLBTC_A,
    HEIGHTTC_A,
    SEX_A,
    HISPALLP_A,
    EDUCP_A,
    PHSTAT_A,
    LSATIS4R_A
  )
View(data_import_R_02_subset)

# Remove missing / nonresponse values using variable-specific codebook values
clean_data_R_02 <- data_import_R_02_subset |>
  filter(
    !AGEP_A %in% c(97, 98, 99),
    !SEX_A %in% c(7, 9),
    !HISPALLP_A %in% c(97, 98, 99),
    !EDUCP_A %in% c(97, 98, 99),
    !PHSTAT_A %in% c(7, 8, 9),
    !LSATIS4R_A %in% c(7, 8, 9),
    !WEIGHTLBTC_A %in% c(996, 997, 998, 999),
    !HEIGHTTC_A %in% c(96, 97, 98, 99)
  ) 
View(clean_data_R_02) # check

# Check cleaned data
summary(clean_data_R_02)
sum(is.na(clean_data_R_02))

# Day 2: Task 2 - Recoding Variables

clean_data_R_02 <- clean_data_R_02 %>%
  mutate(
    EDUCP_A = case_when(
      EDUCP_A >= 0 & EDUCP_A <= 3 ~ 1,
      EDUCP_A == 4 ~ 2,
      EDUCP_A >= 5 & EDUCP_A <= 7 ~ 3,
      EDUCP_A >= 8 & EDUCP_A <= 10 ~ 4
    ),
    EDUCP_A = factor(
      EDUCP_A,
      levels = c(1, 2, 3, 4),
      labels = c(
        "Less than High School",
        "High School Graduate",
        "Some College Education",
        "College Graduate or Better"
      )
    )
  )

# Verify recoding
summary(clean_data_R_02)
table(clean_data_R_02$EDUCP_A)

write_csv(clean_data_R_02, "data/nhis_clean.csv")
```
Interpretation:

```{r}
# Day 3: Task 1: Univariate Analysis & Visualization}

library(tidyverse) #loading tidyverse package 
library(fBasics)

nhis_clean <- read_csv("data/nhis_clean.csv") #loading nhis_clean data into R

"Quantitative Variables"

#AgeP_A (Age)
summary(nhis_clean$AGEP_A) #summary stats for age 
sd(nhis_clean$AGEP_A) #standard deviation for age 

#Base histogram AGEP_A
hist(nhis_clean$AGEP_A, main = "Age Distribution",xlab="Age")

#ggplot histogram AGEP_A
ggplot(nhis_clean,aes(x=AGEP_A)) + #x axis will be Age 
  geom_histogram(binwidth = 5) + #data will be distributed among 5 bins
  xlab("Ages") + #x axis title 
  ggtitle("Age Distribution of Sample Adults") #chart title  

#Base boxplot of AGEP_A
boxplot(nhis_clean$AGEP_A, main = "Age Distribution of Sample Adults") 

#ggplot boxplot of AGEP_A
ggplot(nhis_clean, aes(y=AGEP_A)) + #y axis will be age values 
  geom_boxplot() + #identifying boxplot
  ggtitle("Age Distribution of Sample Adults") + #chart title
  ylab("Ages") #y axis title

 
#WEIGHTLBTC_A (Weight w/out shoes)
summary(nhis_clean$WEIGHTLBTC_A) #Summary stat for weight w/o shoes
sd(nhis_clean$WEIGHTLBTC_A) #standard dev for weight w/o shoes 

basicStats(nhis_clean$WEIGHTLBTC_A) #summary stat for weight w/o shoes 

#Base histogram WEIGHTLBTC_A
hist(nhis_clean$WEIGHTLBTC_A, main = "Weight Distribution of Sample Adults w/out Shoes",xlab="Weight")

#ggplot histogram WEIGHTLBTC_A
ggplot(nhis_clean,aes(x=WEIGHTLBTC_A)) + #x axis will be weight 
  geom_histogram(binwidth = 10) + #data will be distributed among 10 bins
  xlab("Weight") + #x axis title 
  ggtitle("Weight Distribution of Sample Adults w/out Shoes") #chart title  

#Base boxplot of WEIGHTLBTC_A
boxplot(nhis_clean$WEIGHTLBTC_A, main = "Weight Distribution of Sample Adults w/out Shoes") 

#ggplot boxplot of WEIGHTLBTC_A
ggplot(nhis_clean, aes(y=WEIGHTLBTC_A)) + #y axis will be weight values
  geom_boxplot() + #identidying boxplot
  ggtitle("Weight Distribution of Sample Adults w/out Shoes") + #chart title
  ylab("Weight") #y axis title 


#HEIGHTTTC_A (Heigh in Inches)
basicStats(nhis_clean$HEIGHTTC_A) #summary stat for total height in inches  

#Base histogram HEIGHTTTC_A
hist(nhis_clean$HEIGHTTC_A, main = "Height Distribution of Sample Adults",xlab="Total Height in Inches")

#ggplot histogram HEIGHTTTC_A
ggplot(nhis_clean,aes(x=HEIGHTTC_A)) + #x axis will be total height in inches 
  geom_histogram(binwidth = 2) + #data will be distributed among 2 bins
  xlab("Total Height in Inches") + #x axis title 
  ggtitle("Height Distribution of Sample Adults") #chart title  

#Base boxplot of HEIGHTTTC_A
boxplot(nhis_clean$HEIGHTTC_A, main="Total Height Distribution in Inches of Sample Adults") 

#ggplot boxplot of HEIGHTTTC_A
ggplot(nhis_clean, aes(y=HEIGHTTC_A)) + #y axis will be height values
  geom_boxplot() + #identidying boxplot
  ggtitle("Height Distribution of Sample Adults") + #chart title
  ylab("Total Height in Inches") #y axis title 

"Qualitative Variables"

#SEX_A (Gender)

nhis_clean$SEX_A <- factor(nhis_clean$SEX_A, #creating categorical data to a factor 
                           levels = c(1,2), #identifying the values w/in data
                           labels = c("Males","Females")) #relabeling the values w/corresponding label

sex_a_freq_table <- table(nhis_clean$SEX_A) #creating frequency table of sex_a variable 

sex_a_freq_table #printing out the sex_a frequency table 

barplot(sex_a_freq_table, #indicating r base barplot of sex_a variable through freq table 
        main = "Distibution of Sex Among Adult Sample", #chart title
        xlab = "Sex", #x axis title
        ylab = "Frequency", #y axis title
        col = c("lightblue","lightpink")) #bar plot colors 

sex_a_freq_df <- as.data.frame(sex_a_freq_table) #converting freq table to a data frame for ggplot 

ggplot(sex_a_freq_df, aes(x=Var1,y=Freq)) + #using ggplot & the data frame, indicating x as the variables and y as the frequency
  geom_bar(stat = "identity",fill ="purple") + #indicating bar plot
  labs(title = "Distribution of Sex Among Adult Sample", #chart title
       x = "Sex", #x axis title
       y = "Frequency") #y axis title

#HISPALLP_A (Race/Ethnicity)

nhis_clean$HISPALLP_A <- factor(nhis_clean$HISPALLP_A, #creating categorical data to a factor 
                           levels = c(1,2,3,4,5,6,7), #identifying the values w/in data
                           labels = c("Hispanic","NHWhite","NHAA","NHAsian Only","NHAIAN","NHAIAN&Other","Other")) #relabeling the values w/corresponding label

hispallp_a_freq_table <- table(nhis_clean$HISPALLP_A) #creating frequency table of HISPALLPa variable 

hispallp_a_freq_table #printing out the hispallp_a frequency table 

barplot(hispallp_a_freq_table, #indicating r base barplot of hispallp_a variable through freq table 
        main = "Distibution of Race/Ethnicity Among Adult Sample", #chart title
        xlab = "Race/Ethniticy", #x axis title
        ylab = "Frequency", #y axis title
        col = terrain.colors(7), #bar plot colors 
        cex.names = 0.6) #label size

hispallp_a_freq_df <- as.data.frame(hispallp_a_freq_table) #converting freq table to a data frame for ggplot 

ggplot(hispallp_a_freq_df, aes(x=Var1,y=Freq)) + #using ggplot & the data frame, indicating x as the variables and y as the frequency
  geom_bar(stat = "identity",fill ="lightgreen") + #indicating bar plot
  labs(title = "Distribution of Race/Ethnicity Among Adult Sample", #chart title
       x = "Race/Ethnicity", #x axis title
       y = "Frequency") #y axis title

#EDUCP_A (Education Level)

"EDCUP_A was already recoded and turned into a factor in previous task"

educp_a_freq_table <- table(nhis_clean$EDUCP_A) #creating frequency table of EDUCP_A variable 

educp_a_freq_table #printing out the educp_a frequency table 

barplot(educp_a_freq_table, #indicating r base barplot of educp_a variable through freq table 
        main = "Distibution of Education Level Among Adult Sample", #chart title
        xlab = "Education Level", #x axis title
        ylab = "Frequency", #y axis title
        col = rainbow(4), #bar plot colors 
        cex.names = 0.45) #label size

educp_a_freq_df <- as.data.frame(educp_a_freq_table) #converting freq table to a data frame for ggplot 

ggplot(educp_a_freq_df, aes(x=Var1,y=Freq)) + #using ggplot & the data frame, indicating x as the variables and y as the frequency
  geom_bar(stat = "identity",fill ="blue") + #indicating bar plot
  labs(title = "Distribution of Education Level Among Adult Sample", #chart title
       x = "Race/Ethnicity", #x axis title
       y = "Frequency") + #yaxis title 
  theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1)) #adjusts labels on x axis

#PHSTAT_A (General Health)

nhis_clean$PHSTAT_A<- factor(nhis_clean$PHSTAT_A, #creating categorical data to a factor 
                                levels = c(1,2,3,4,5), #identifying the values w/in data
                                labels = c("Excellent","Very Good","Good","Fair","Poor")) #relabeling the values w/corresponding label

phstat_a_freq_table <- table(nhis_clean$PHSTAT_A) #creating frequency table of PHSTAT_A variable 

phstat_a_freq_table #printing out the phstat_a frequency table 

barplot(phstat_a_freq_table, #indicating r base barplot of phstat_a variable through freq table 
        main = "Distibution of Health Status Among Adult Sample", #chart title
        xlab = "General Health Status", #x axis title
        ylab = "Frequency", #y axis title
        col = topo.colors(5), #bar plot colors 
        cex.names = 0.8) #label size

phstat_a_freq_df <- as.data.frame(phstat_a_freq_table) #converting freq table to a data frame for ggplot 

ggplot(phstat_a_freq_df, aes(x=Var1,y=Freq)) + #using ggplot & the data frame, indicating x as the variables and y as the frequency
  geom_bar(stat = "identity",fill ="gray") + #indicating bar plot
  labs(title = "Distribution of Health Status Among Adult Sample", #chart title
       x = "General Health Status", #x axis title
       y = "Frequency") #y axis title

#LSATIS4R_A (Quality of Life)

nhis_clean$LSATIS4R_A<- factor(nhis_clean$LSATIS4R_A, #creating categorical data to a factor 
                             levels = c(1,2,3,4), #identifying the values w/in data
                             labels = c("Very Satisfied","Satisfied","Dissatisfied","Very Dissatisfied")) #relabeling the values w/corresponding label

lsatis4r_a_freq_table <- table(nhis_clean$LSATIS4R_A) #creating frequency table of LSATIS4R_A variable 

lsatis4r_a_freq_table #printing out the lsatis4r_a frequency table 

barplot(lsatis4r_a_freq_table, #indicating r base barplot of lsatis4r_a variable through freq table 
        main = "Distibution of Quality of Life Among Adult Sample", #chart title
        xlab = "Quality of Life", #x axis title
        ylab = "Frequency", #y axis title
        col = heat.colors(4), #bar plot colors 
        cex.names = 0.8) #label size

lstatis4r_a_freq_df <- as.data.frame(lsatis4r_a_freq_table) #converting freq table to a data frame for ggplot 

ggplot(lstatis4r_a_freq_df, aes(x=Var1,y=Freq)) + #using ggplot & the data frame, indicating x as the variables and y as the frequency
  geom_bar(stat = "identity",fill = "orange") + #indicating bar plot
  labs(title = "Distribution of Quality of Life Among Adult Sample", #chart title
       x = "Quality of Life", #x axis title
       y = "Frequency") #y axis title
```
Interpretation:



```{r}
# Day 3: Task 2: Bivariate Analysis (Boxplots Age v Sex & Age v Education)}

"Age v Sex Boxplot"
boxplot(AGEP_A~SEX_A, data = nhis_clean,#boxplot
        main = "Boxplot of Age by Sex ", #chart title
        xlab = "Sex", #x axis title
        ylab = "Age") #y axis title

"Age v Education Boxplot"
boxplot(AGEP_A~EDUCP_A, data = nhis_clean,#boxplot
        main = "Boxplot of Age by Education ", #chart title
        xlab = "Education Level", #x axis title
        ylab = "Age") #y axis title

"Age v Sex GGPlot"
ggplot(nhis_clean,aes(x=SEX_A,y=AGEP_A)) + #indicating gglot, data set, x & y axis
  geom_boxplot(aes(fill = factor(SEX_A))) + #identifying boxplot chart & filling box with Sex values
  labs(title = "Boxplot of Age by Sex", #chart title
       x = "Sex", #x axis title
       y = "Age", #y axis title
       fill = "Sex") #title of legend 

"Age v Education ggplot"
ggplot(nhis_clean,aes(x=EDUCP_A,y=AGEP_A)) + #indicating gglot, data set, x & y axis
  geom_boxplot(aes(fill = factor(EDUCP_A))) + #identifying boxplot chart & filling box with Educ values
  labs(title = "Boxplot of Age by Education Level", #chart title
       x = "Education Level", #x axis title
       y = "Age", #y axis title
       fill = "Education Level") #title of legend 

#Clustered Bar Chart (General Health & Life Statisfaction)

ggplot(nhis_clean,aes(x=PHSTAT_A,fill = LSATIS4R_A)) + #identifying ggplot, x axis and y axis 
  geom_bar(position = "dodge") + #indicating bar chart 
  labs(title = "Relationship between General Health & Life Satisfaction", #chart title
       x = "General Health Status", #x axis title
       y = "Count", #y axis title
       fill = "Life Satisfaction") #legend title 

#Scatterplot of Height vs Weight 

plot(nhis_clean$HEIGHTTC_A,nhis_clean$WEIGHTLBTC_A, #scatterplot usingt plot() for height vs weight 
     main = "Height(in.) vs Weight (lbs.) Scatterplot", #chart title
     xlab = "Height (in.)", #x axis title
     ylab = "Weight(lbs.)") #y axis title

ggplot(nhis_clean,aes(x=HEIGHTTC_A,y=WEIGHTLBTC_A)) + #using ggplot for scatterplot, indicating x and y values 
  geom_point() + #indicating points for each observation 
  labs(title = "Height(in.) vs Weight (lbs.) Scatterplot", #chart title
       x = "Height (in.)", #x axis title
       y = "Weight(lbs.)") #y axis title

cor(nhis_clean$HEIGHTTC_A,nhis_clean$WEIGHTLBTC_A) #calculating & reporting correlation coefficient; results 0.5158, medium correlation 
```
Interpretation:



```{r}
# Day 4 Task 1: Multivariate_Visualization

# Enhanced plot by General Health Status (PHSTAT_A)
ggplot(nhis_clean, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = factor(SEX_A))) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ PHSTAT_A) +
  labs(
    title = "Height vs Weight by Sex and General Health Status",
    x = "Height (in.)",
    y = "Weight (lbs)",
    color = "Sex"
  ) +
  theme_linedraw()

# or Enhanced plot by Education Level (PHSTAT_A)
ggplot(nhis_clean, aes(x = HEIGHTTC_A, y = WEIGHTLBTC_A, color = factor(SEX_A))) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ EDUCP_A) +
  labs(
    title = "Height vs Weight by Sex and Education Level",
    x = "Height (in.)",
    y = "Weight (lbs)",
    color = "Sex"
  ) +
  theme_linedraw()

# Create a correlation scatter plot matrix using pairs.panels() from the psych library to visualize AGEP_A, WEIGHTLBTC_A, and HEIGHTTC_A

# Install psych 
install.packages("psych")
# Load the package
library(psych)

#Isolate three variables 
nhis_pp_variables <- nhis_clean[, c("AGEP_A", "WEIGHTLBTC_A", "HEIGHTTC_A")] 

# Create the pairs.panels correlation plot
pairs.panels(
  nhis_pp_variables,
  method = "pearson",     # correlation method
  hist.col = "chartreuse4", 
  density = TRUE,         # add density plots
  ellipses = TRUE         # add correlation ellipses
)
```
Interpretation:

The multivariate visualizations provide insight into how height and weight vary across sex and different social and health-related groupings.

Height vs Weight by General Health Status:
The scatterplot shows a pretty clear upward trend—taller people generally weigh more across every health group (excellent through poor). That relationship stays pretty consistent no matter the category of health. “Fair” and “Poor” health groups have a wider spread in weight, meaning body size is more varied when compared to the tighter clustering you see in better health categories. Across all the panels, males consistently sit at higher weight values than females at similar heights, which keeps showing up as a pretty stable sex difference in body size.

Height vs Weight by Education Level:
The same pattern is seen here too from above as height and weight move together across all education groups. People with higher education (college graduates and above) look a bit more tightly clustered, while lower education groups show more variation in weight at similar heights. But the overall relationship doesn’t really change by education level. Males tend to cluster higher in weight than females across every group, furthering proving that the pattern stays consistent no matter the education category.

Correlation Matrix (Age, Height, Weight):
The correlation results back up what the plots are showing. Height and weight have a moderate positive relationship (around 0.50), meaning taller people tend to weigh more. Age, on the other hand, barely relates to either variable—almost zero with height (-0.02) and very weak with weight (-0.09). As a resut it can be said that in this dataset, body size doesn’t really shift with age. The histograms also show age is pretty spread out across the sample, while height and weight are more normally concentrated around their middle ranges.
